---
title: "Omics: Functional Gene Annotation Analysis"
author: "Seunghyeon Kang"
date: "2026-05-09"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This analysis explores functional gene annotations across four bacterial
strain types: Mal, Dimeth, Chlorp, and Diaz. Protein annotations were
obtained from OmicsBox and are compared to identify shared and unique
functional categories across strains.

---

## Setup

```{r}
#| label: load-packages
#| message: false

# Load all required packages
library(tidyverse)
library(here)
```

```{r}
#| label: define-colors

# Custom colorblind-friendly palette used throughout visualizations
colors_van_likes <- c(
  "#E69F00", # orange
  "#009E73", # bluish green
  "#F0E442", # yellow
  "#D55E00", # vermillion
  "#0072B2", # blue
  "#CC79A7"  # purple/pink
)
```

---

## Data Cleaning

```{r}
#| label: helper-function

# Helper: read a list of OmicsBox .txt files and add strain metadata
read_omics_files <- function(files, strain_type) {
  files |>
    set_names() |>
    map_dfr(
      ~ read_tsv(.x, show_col_types = FALSE),
      .id = "file"
    ) |>
    mutate(
      Strain      = str_remove(file, "_omicsbox_table\\.txt$"),
      StrainType  = strain_type,
      Description = str_remove(Description, "^MULTISPECIES:\\s*")
    ) |>
    select(-file)
}
```

```{r}
#| label: load-data

# --- Mal / Dimeth strains (already combined in MalDimeth.csv) ---
mal_dimeth_df <- read_csv(
  here("MalDimeth.csv"),
  show_col_types = FALSE
) |>
  mutate(
    StrainType = case_when(
      str_detect(Strain, "^Dimeth") ~ "Dimeth",
      str_detect(Strain, "^Mal")   ~ "Mal",
      TRUE                          ~ "Other"
    ),
    Description = str_remove(Description, "^MULTISPECIES:\\s*")
  )

# --- Chlorp strains ---
chlorp_files <- c(
  "Chlorp2_omicsbox_table.txt",
  "Chlorp5_omicsbox_table.txt",
  "Chlorp7_omicsbox_table.txt",
  "Chlorp8_omicsbox_table.txt",
  "Chlorp10_omicsbox_table.txt",
  "Chlorp11_omicsbox_table.txt"
)

# --- Diaz strains ---
diaz_files <- c(
  "Diaz11_omicsbox_table.txt",
  "Diaz23_omicsbox_table.txt",
  "Diaz26_omicsbox_table.txt",
  "Diaz28_omicsbox_table.txt"
)

chlorp_df <- read_omics_files(chlorp_files, "Chlorp")
diaz_df   <- read_omics_files(diaz_files,   "Diaz")
```

```{r}
#| label: combine-and-categorize

# Merge all strain data into one dataframe and save clean version
df <- bind_rows(mal_dimeth_df, chlorp_df, diaz_df)

write_csv(df, here("data_clean", "MalChlorpDimethDiaz.csv"))

# Helper: categorize gene annotations into broad functional groups
categorize_function <- function(desc) {
  desc <- str_to_lower(desc)
  case_when(
    str_detect(
      desc,
      "transporter|permease|abc|symporter|receptor|efflux"
    ) ~ "Transport",
    str_detect(
      desc,
      "regulator|transcription|sigma|response regulator|kinase|phosphatase|sensor"
    ) ~ "Regulation",
    str_detect(
      desc,
      "dehydrogenase|oxidoreductase|synthase|ligase|transferase|hydrolase|metabolic|biosynthetic|catabolic"
    ) ~ "Metabolism",
    str_detect(
      desc,
      "stress|resistance|glutathione|repair|tellurite|osm|detox"
    ) ~ "Stress/Defense",
    str_detect(
      desc,
      "ribosomal|elongation factor|polymerase|trna ligase"
    ) ~ "Information Processing",
    TRUE ~ "Other"
  )
}
```

---

## Visualizations

### Functional gene categories by strain

```{r}
#| label: fig-categories-by-strain
#| fig-cap: "Stacked bar chart of annotated gene counts per strain, grouped
#|   by broad functional category."
#| fig-alt: "Stacked bar chart showing the number of annotated genes across
#|   individual bacterial strains. Each bar is divided into colored segments
#|   representing functional categories: Transport, Metabolism, Regulation,
#|   Stress/Defense, Information Processing, and Other. Transport and
#|   Metabolism are the most common categories across all strains."

# Count genes per functional category within each strain
df |>
  mutate(Category = categorize_function(Description)) |>
  count(Strain, Category) |>
  ggplot(aes(x = Strain, y = n, fill = Category)) +
  geom_col() +
  scale_fill_manual(values = colors_van_likes) +
  labs(
    title   = "Functional gene categories across strains",
    x       = "Strain",
    y       = "Number of annotated genes",
    fill    = "Functional category",
    caption = "Source: OmicsBox functional annotation tables"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Functional annotations were grouped into broad categories including
metabolism, transport, regulation, stress/defense, and information
processing. Transport- and metabolism-related genes were abundant across
all Mal strains, with smaller differences among strains in stress/defense
and regulatory genes.

---

### Gene sharing within strain types

```{r}
#| label: fig-gene-sharing-within-type
#| fig-cap: "Distribution of how many strains within each strain type share
#|   a given gene annotation. High counts at x = 1 indicate many
#|   strain-specific genes; high counts at the maximum x value indicate
#|   a large shared core genome."
#| fig-alt: "Faceted bar charts, one panel per strain type (Mal, Dimeth,
#|   Chlorp, Diaz), showing how many gene annotations are shared across
#|   1, 2, 3, or more strains within each type. Mal strains show many
#|   annotations shared across all strains, while Dimeth shows higher
#|   proportions of strain-unique annotations."

# Count how many strains within a type share each annotation
shared_genes <- df |>
  filter(!is.na(Description), Description != "") |>
  distinct(StrainType, Strain, Description) |>
  count(StrainType, Description, name = "n_strains") |>
  count(StrainType, n_strains)

ggplot(shared_genes, aes(x = factor(n_strains), y = n, fill = factor(n_strains))) +
  geom_col() +
  scale_fill_viridis_d() +
  facet_wrap(~ StrainType, scales = "free_y") +
  labs(
    title   = "Distribution of shared and unique gene annotations within strain types",
    x       = "Number of strains sharing a gene annotation",
    y       = "Number of gene annotations",
    fill    = "# of strains",
    caption = "Source: OmicsBox functional annotation tables"
  ) +
  theme_classic()
```

The distribution of gene annotations shows a high number of genes shared
across all strains and many genes unique to individual strains. Genes
present in all strains likely represent core functions essential for
survival, while strain-specific genes suggest functional diversity and
adaptation to different environmental conditions.

Mal strains show a larger proportion of genes shared across all or most
strains, indicating a more conserved core genome. In contrast, Dimeth
strains have a higher proportion of genes unique to individual strains
and fewer genes shared across all strains, suggesting greater genomic
variability.

---

### Gene annotation sharing across strain types

```{r}
#| label: fig-sharing-across-types
#| fig-cap: "Proportional stacked bar chart showing what fraction of each
#|   strain type's gene annotations are unique to that type, shared with
#|   all four types, or partially shared with 2–3 types."
#| fig-alt: "Stacked proportional bar chart with one bar per strain type.
#|   Each bar is divided into three segments: Unique (annotation found only
#|   in that strain type), Shared (All) (present in all four strain types),
#|   and Partial (present in 2 or 3 strain types). All strain types show a
#|   mix of unique and shared annotations."

# Classify annotations as unique, partially shared, or shared across all types
n_types_total <- df |>
  distinct(StrainType) |>
  nrow()

df |>
  filter(!is.na(Description), Description != "") |>
  distinct(StrainType, Description) |>
  add_count(Description, name = "n_types") |>
  mutate(
    sharing_type = case_when(
      n_types == 1            ~ "Unique",
      n_types == n_types_total ~ "Shared (All)",
      TRUE                    ~ "Partial"
    )
  ) |>
  count(StrainType, sharing_type) |>
  ggplot(aes(x = StrainType, y = n, fill = sharing_type)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = colors_van_likes[c(1, 5, 2)]) +
  labs(
    title   = "Gene annotation sharing across strain types",
    x       = "Strain type",
    y       = "Proportion of gene annotations",
    fill    = "Sharing type",
    caption = paste(
      "Unique: found in 1 strain type only.",
      "Shared (All): present in all 4 strain types.",
      "Partial: present in 2–3 strain types.",
      sep = " "
    )
  ) +
  theme_classic()
```

Gene annotations were compared based on functional descriptions, allowing
identification of shared and unique functional categories across strain
types.

---

### Functional composition by strain type

```{r}
#| label: fig-composition-by-type
#| fig-cap: "Proportional stacked bar chart of functional gene categories
#|   aggregated by strain type, allowing comparison of overall functional
#|   composition between Mal, Dimeth, Chlorp, and Diaz."
#| fig-alt: "Proportional stacked bar chart with one bar per strain type.
#|   Each bar shows the relative proportion of functional categories
#|   (Transport, Metabolism, Regulation, Stress/Defense, Information
#|   Processing, Other) within that strain type. The proportions are broadly
#|   similar across types, with Transport and Metabolism dominant."

# Summarise functional category proportions at the strain-type level
df |>
  mutate(Category = categorize_function(Description)) |>
  count(StrainType, Category) |>
  ggplot(aes(x = StrainType, y = n, fill = Category)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = colors_van_likes) +
  labs(
    title   = "Functional composition by strain type",
    x       = "Strain type",
    y       = "Proportion of genes",
    fill    = "Functional category",
    caption = "Source: OmicsBox functional annotation tables"
  ) +
  theme_classic()
```

Functional proportions are broadly similar across strain types, with
Transport and Metabolism dominating in all groups. Minor differences in
Stress/Defense and Regulation proportions may reflect adaptation to
distinct ecological niches.
