---
title: "Omics"
author: "Van Stewart"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
colorsVanLikes <- c(
  
  "#E69F00", # orange
  "#009E73", # bluish green
  "#F0E442",  # yellow
  "#D55E00", # vermillion
  "#0072B2", # blue
  "#CC79A7" # purple/pink
  
)


```

```{r}
library(tidyverse)

mal_dimeth_df <- read_csv("MalDimeth.csv") |>
  mutate(
    StrainType = case_when(
      str_detect(Strain, "^Dimeth") ~ "Dimeth",
      str_detect(Strain, "^Mal") ~ "Mal",
      TRUE ~ "Other"
    )
  )

chlorp_files <- c(
  "Chlorp2_omicsbox_table.txt",
  "Chlorp5_omicsbox_table.txt",
  "Chlorp7_omicsbox_table.txt",
  "Chlorp10_omicsbox_table.txt",
  "Chlorp11_omicsbox_table.txt",
  "Chlorp8_omicsbox_table.txt"
)

diaz_files <- c(
  "Diaz11_omicsbox_table.txt",
  "Diaz23_omicsbox_table.txt",
  "Diaz26_omicsbox_table.txt",
  "Diaz28_omicsbox_table.txt"
)

diaz_df <- diaz_files %>%
  set_names() %>%
  map_dfr(~ read_tsv(.x, show_col_types = FALSE), .id = "File") %>%
  mutate(Strain = str_remove(File, "_omicsbox_table\\.txt$"),
    StrainType = "Diaz",
    Description = str_remove(Description, "^MULTISPECIES:\\s*")
  )


chlorp_df <- chlorp_files %>%
  set_names() %>%
  map_dfr(~ read_tsv(.x, show_col_types = FALSE), .id = "File") %>%
  mutate(
    Strain = str_remove(File, "_omicsbox_table\\.txt$"),
    StrainType = "Chlorp",
    Description = str_remove(Description, "^MULTISPECIES:\\s*")
  )

df <- bind_rows(mal_dimeth_df, chlorp_df, diaz_df)|>
  write_csv("MalChlorpDimethDiaz.csv")

categorize_function <- function(desc) {
  desc <- str_to_lower(desc)
  case_when(
    str_detect(desc, "transporter|permease|abc|symporter|receptor|efflux") ~ "Transport",
    str_detect(desc, "regulator|transcription|sigma|response regulator|kinase|phosphatase|sensor") ~ "Regulation",
    str_detect(desc, "dehydrogenase|oxidoreductase|synthase|ligase|transferase|hydrolase|metabolic|biosynthetic|catabolic") ~ "Metabolism",
    str_detect(desc, "stress|resistance|glutathione|repair|tellurite|osm|detox") ~ "Stress/Defense",
    str_detect(desc, "ribosomal|elongation factor|polymerase|trna ligase") ~ "Information Processing",
    TRUE ~ "Other"
  )
}

```


.
```{r}


df|>
  mutate(Category = categorize_function(Description))|>
  count(Strain, Category)|>
  ggplot(aes(x = Strain, y = n, fill = Category))+
  scale_fill_manual(values = colorsVanLikes)+
  geom_col()+
  labs(
    title = "Functional gene categories across strains",
    x = "Strain",
    y = "Number of annotated genes"
  )+
  theme_classic()+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
> Functional annotations were grouped into broad categories including metabolism, transport, regulation, stress/defense, and information processing. Transport- and metabolism-related genes were abundant across all Mal strains, with smaller differences among strains in stress/defense and regulatory genes

```{r}


shared_genes <- df |>
  # remove missing descriptions
  filter(!is.na(Description), Description != "") |>
  
  # one gene per strain
  distinct(StrainType, Strain, Description) |>
  
  
  count(StrainType, 
        Description, name = "n_strains") |>
  
  # distribution of sharing
  count(StrainType, n_strains)

ggplot(shared_genes, aes(x = factor(n_strains), y = n, fill = factor(n_strains))) +
  geom_col() +
  scale_fill_viridis_d() +
  facet_wrap(~ StrainType, scales = "free_y") +
  labs(
    title = "Distribution of shared and unique gene annotations within strain types",
    x = "Number of strains within a strain type containing a gene annotation",
    y = "Number of gene annotations",
    fill = "# of strains"
  ) +
  theme_classic()
```
> The distribution of gene annotations shows a high number of genes shared across all strains and many genes unique to individual strains. Genes present in all strains likely represent core functions essential for survival, while strain-specific genes suggest functional diversity and adaptation to different environmental conditions.

> Mal strains show a larger proportion of genes shared across all or most strains (n ≈ total strains), indicating a more conserved core genome. In contrast, Dimeth strains have a higher proportion of genes unique to individual strains and fewer genes shared across all strains, suggesting greater genomic variability.

```{r}

n_types_total <- df %>%
  distinct(StrainType) %>%
  nrow()

df |>
  filter(!is.na(Description), Description != "") |>
  distinct(StrainType, Description) |>
  add_count(Description, name = "n_types") |>
  mutate(Type = case_when(
    n_types == 1 ~ "Unique",
    n_types == n_types_total ~ "Shared (All)",
    TRUE ~ "Partial"
  )) |>
  count(StrainType, Type) |>
  ggplot(aes(x = StrainType, y = n, fill = Type)) +
  geom_col(position = "fill") +
  labs(
    title = "Gene annotation sharing across strain types",
    x = "Strain type",
    y = "Proportion of gene annotations"
  ) +
  theme_classic()
```
> Unique: only in 1 strain type
> Shared (All): present in all 4 strain types 
> Partial: in 2–3 types
> Gene annotations were compared based on functional descriptions, allowing identification of shared and unique functional categories across strain types.




```{r}
df |>
  mutate(Category = categorize_function(Description)) |>
  count(StrainType, Category) |>
  ggplot(aes(x = StrainType, y = n, fill = Category)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = colorsVanLikes) +
  labs(y = "Proportion of genes") +
  theme_classic()


```


