---
title: "Portfolio Part 2: Data Cleaning and Rough Draft"
author: "Seunghyeon Kang"
format: html
editor: visual
execute:
  echo: true
  warning: false
---

# Research Question

This project explores functional differences across pesticide-degrading bacterial strains.

-   What functional gene categories are most abundant across strain types?
-   Do different strains show similar or different functional profiles?

# Data Cleaning

```{r}
library(tidyverse)
library(stringr)
library(purrr)

colorsVanLikes <- c(
  "#E69F00", "#009E73", "#F0E442",
  "#D55E00", "#0072B2", "#CC79A7"
)
```

#Load and Combine Data

```{r}
# Read the Mal and Dimeth dataset
mal_dimeth_df <- read_csv("MalDimeth.csv", show_col_types = FALSE) %>%
  mutate(
    StrainType = case_when(
      str_detect(Strain, "^Dimeth") ~ "Dimeth",
      str_detect(Strain, "^Mal") ~ "Mal",
      TRUE ~ "Other"
    )
  )

# List Chlorp annotation files
chlorp_files <- c(
  "Chlorp2_omicsbox_table.txt",
  "Chlorp5_omicsbox_table.txt",
  "Chlorp7_omicsbox_table.txt",
  "Chlorp10_omicsbox_table.txt",
  "Chlorp11_omicsbox_table.txt",
  "Chlorp8_omicsbox_table.txt"
)

# List Diaz annotation files
diaz_files <- c(
  "Diaz11_omicsbox_table.txt",
  "Diaz23_omicsbox_table.txt",
  "Diaz26_omicsbox_table.txt",
  "Diaz28_omicsbox_table.txt"
)

# Read and clean Diaz files
diaz_df <- diaz_files %>%
  set_names() %>%
  map_dfr(~ read_tsv(.x, show_col_types = FALSE), .id = "File") %>%
  mutate(
    Strain = str_remove(File, "_omicsbox_table\\.txt$"),
    StrainType = "Diaz",
    Description = str_remove(Description, "^MULTISPECIES:\\s*")
  )

# Read and clean Chlorp files
chlorp_df <- chlorp_files %>%
  set_names() %>%
  map_dfr(~ read_tsv(.x, show_col_types = FALSE), .id = "File") %>%
  mutate(
    Strain = str_remove(File, "_omicsbox_table\\.txt$"),
    StrainType = "Chlorp",
    Description = str_remove(Description, "^MULTISPECIES:\\s*")
  )

# Combine all annotation tables into one dataset
df <- bind_rows(mal_dimeth_df, chlorp_df, diaz_df)
```

#Clean and Prepare Data

```{r}
# Remove blank descriptions and convert them to NA
df_clean <- df %>%
  mutate(Description = na_if(Description, "")) %>%
  filter(!is.na(Description))

# Keep one unique annotation per strain
df_clean <- df_clean %>%
  distinct(StrainType, Strain, Description, .keep_all = TRUE)
```

#Create Functional Categories

```{r}
# Assign broad functional categories based on keywords in Description
categorize_function <- function(desc) {
  desc <- str_to_lower(desc)
  case_when(
    str_detect(desc, "transporter|permease|abc|symporter|receptor|efflux") ~ "Transport",
    str_detect(desc, "regulator|transcription|sigma|response regulator|kinase|sensor") ~ "Regulation",
    str_detect(desc, "dehydrogenase|oxidoreductase|synthase|ligase|transferase|hydrolase|metabolic") ~ "Metabolism",
    str_detect(desc, "stress|resistance|repair|detox|glutathione") ~ "Stress/Defense",
    str_detect(desc, "ribosomal|polymerase|trna|elongation") ~ "Information Processing",
    TRUE ~ "Other"
  )
}

df_clean <- df_clean %>%
  mutate(Category = categorize_function(Description))
```

#Save Clean Data

```{r}
dir.create("data_clean", showWarnings = FALSE)
write_csv(df_clean, "data_clean/omics_clean_data.csv")
```

#Rough Draft Visualization

```{r}
df_clean %>%
  count(Strain, Category) %>%
  ggplot(aes(x = Strain, y = n, fill = Category)) +
  geom_col() +
  scale_fill_manual(values = colorsVanLikes) +
  labs(
    title = "Functional Gene Categories Across Strains",
    x = "Strain",
    y = "Number of Genes"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#Interpretation

This plot shows the distribution of broad functional gene categories across bacterial strains. Metabolism and transport-related genes appear to be the most abundant across many strains, suggesting that these functions are common among pesticide-degrading bacteria. At the same time, smaller differences across strains may indicate variation in functional specialization.

#What I Want to Improve

For the final version of this project, I plan to add additional visualizations that show how gene annotations are shared across strains and strain types. I also want to improve the styling and presentation of the graphs for my website.
